Tracing the evolution of short-period binaries with super-synchronous fast-rotators

### 2024-01-15 src/tex/ms.tex
accepted version of the paper

% Define document class
\documentclass[twocolumn]{aastex631}
\usepackage{showyourwork}

% Begin!
\begin{document}

% Title
\title{An open source scientific article}

% Author list
\author{@NikolayBritavskiyAstro}

% Abstract with filler text
\begin{abstract}
    Lorem ipsum dolor sit amet, consectetuer adipiscing elit.
    Ut purus elit, vestibulum ut, placerat ac, adipiscing vitae, felis.
    Curabitur dictum gravida mauris, consectetuer id, vulputate a, magna.
    Donec vehicula augue eu neque, morbi tristique senectus et netus et.
    Mauris ut leo, cras viverra metus rhoncus sem, nulla et lectus vestibulum.
    Phasellus eu tellus sit amet tortor gravida placerat.
    Integer sapien est, iaculis in, pretium quis, viverra ac, nunc.
    Praesent eget sem vel leo ultrices bibendum.
    Aenean faucibus, morbi dolor nulla, malesuada eu, pulvinar at, mollis ac.
    Curabitur auctor semper nulla donec varius orci eget risus.
    Duis nibh mi, congue eu, accumsan eleifend, sagittis quis, diam.
    Duis eget orci sit amet orci dignissim rutrum.
\end{abstract}

% Main body with filler text
\section{Introduction}
\label{sec:intro}

Lorem ipsum dolor sit amet, consectetuer adipiscing elit.
Ut purus elit, vestibulum ut, placerat ac, adipiscing vitae, felis.
Curabitur dictum gravida mauris, consectetuer id, vulputate a, magna.
Donec vehicula augue eu neque, morbi tristique senectus et netus et.
Mauris ut leo, cras viverra metus rhoncus sem, nulla et lectus vestibulum.
Phasellus eu tellus sit amet tortor gravida placerat.
Integer sapien est, iaculis in, pretium quis, viverra ac, nunc.
Praesent eget sem vel leo ultrices bibendum.
Aenean faucibus, morbi dolor nulla, malesuada eu, pulvinar at, mollis ac.
Curabitur auctor semper nulla donec varius orci eget risus.
Duis nibh mi, congue eu, accumsan eleifend, sagittis quis, diam.
Duis eget orci sit amet orci dignissim rutrum.

Nam dui ligula, fringilla a, euismod sodales, sollici- tudin vel, wisi.
Morbi auctor lorem non justo, nam lacus libero, pretium at, lobortis vitae.
Donec aliquet, tortor sed accumsan bibendum, erat ligula aliquet magna.
Morbi ac orci et nisl hendrerit mollis, suspendisse ut massa, cras nec ante.
Pellentesque a nulla cum sociis natoque penatibus et magnis dis parturient.
Aliquam tincidunt urna, nulla ullamcorper vestibulum turpis.
Pellentesque cursus luctus mauris \citep{Luger2021}.

\bibliography{bib}

>>>>

% aa.dem
% AA vers. 8.2, LaTeX class for Astronomy & Astrophysics
% demonstration file
% (c) EDP Sciences
% -----------------------------------------------------------------------
%
% \documentclass[referee]{aa} % for a referee version
% \documentclass[onecolumn]{aa} % for a paper on 1 column
% \documentclass[longauth]{aa} % for the long lists of affiliations
% \documentclass[rnote]{aa} % for the research notes
% \documentclass[letter]{aa} % for the letters
% \documentclass[bibyear]{aa} % if the references are not structured
% according to the author-year natbib style

%
\documentclass{aa}

%
\usepackage{graphicx}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{txfonts}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{xcolor}
% \usepackage{fontspec}
\usepackage{amsmath}
\usepackage{orcidlink}
\usepackage[normalem]{ulem}
\newcommand{\kms}{$\mathrm{km\,s^{-1}}$}
\newcommand{\vsini} {$v\sin\,i$}
\DeclareRobustCommand{\Eqref}[1]{Equation~\ref{#1}}
\DeclareRobustCommand{\Figref}[1]{Figure~\ref{#1}}
\DeclareRobustCommand{\Tabref}[1]{Table~\ref{#1}}
\DeclareRobustCommand{\Secref}[1]{Section~\ref{#1}}
\DeclareRobustCommand{\Appref}[1]{Appendix~\ref{#1}}
\newcommand{\todo}[1]{{\large
    $\blacksquare$~\textbf{\color{red}[#1]}~$\blacksquare$}}
\newcommand{\mr}[1]{\textcolor{green!50!black}{#1}}
\newcommand{\nb}[1]{\textcolor{blue}{#1}}
\usepackage{showyourwork}

\begin{document}

\graphicspath{{./figures/}}

\title{Tracing the evolution of short-period binaries with super-synchronous fast-rotators}

\author{N. Britavskiy\inst{1,2}\orcidlink{0000-0003-3996-0175},
  M. Renzo\inst{3,4}\orcidlink{0000-0002-6718-9472},
  Y.~Naz\'e\inst{1}\thanks{FNRS Senior Research Associate}\orcidlink{0000-0003-4071-9346},
  G.~Rauw\inst{1}\orcidlink{0000-0003-4715-9871},
  P.~Vynatheya\inst{5}\orcidlink{0000-0003-3736-2059}
}

\institute{Universit\'e de Li\`ege, Quartier Agora (B5c, Institut d’Astrophysique et de Geophysique), All\'ee du 6 Ao\"ut 19c, B-4000 Sart Tilman, Li\`ege, Belgium \and Royal Observatory of Belgium, Avenue Circulaire/Ringlaan 3, B-1180 Brussels, Belgium
\and Steward Observatory, University of Arizona, 933 N. Cherry Avenue, Tucson, AZ 85721, USA \and
  Center for Computational Astrophysics, Flatiron Institute, New York, NY 10010, USA \and
  Max-Planck-Institut für Astrophysik, Karl-Schwarzschild-Strasse 1, 85748 Garching bei München, Germany
}
\offprints{mbritavskiy@uliege.be}

\date{Submitted/Accepted}

\titlerunning{Fast-rotating massive stars in a short-period binaries}

\authorrunning{N. Britavskiy et al.}

% \abstract{}{}{}{}{}
% 5 {} token are mandatory

\abstract
% context heading (optional)
% {} leave it empty if necessary
{The initial distribution of rotational velocities of stars is still
  poorly known, and how the stellar spin evolves from birth to the
  various end-points of stellar evolution is an actively debated
  topic. Binary interactions are often invoked to explain the
  existence of extremely fast-rotating stars (\vsini~
  $\gtrsim 200$\,\kms). The primary mechanisms through which binaries
  can spin up stars are tidal interactions, mass-transfer, and
  possibly mergers. However, fast rotation could also be primordial,
  that is a result of the star formation process. 
  To evaluate these scenarios, we investigate in detail the
  evolution of three known fast-rotating stars in short-period spectroscopic and eclipsing
  binaries, namely HD\,25631, HD\,191495, and HD\,46485, with
  primaries of masses of 7, 15, and 24 $M_{\odot}$ respectively, with
  companions of $\sim1\,M_\odot$ and orbital periods of less than 7 days. These systems belong to a
  recently identified class of binaries with extreme mass ratios, whose evolutionary origin is still poorly understood.}
% aims heading (mandatory)
{We evaluate in detail three scenarios that could explain the fast
    rotation observed in these binaries: it could be primordial, a
    product of mass transfer, or the result of a merger within an
    originally triple system. We also discuss the future evolution of
    these systems to shed light on the impact of fast rotation on
    binary products.}
% methods heading (mandatory)
  {We compute grids of single and binary MESA models varying tidal
      forces and initial binary architectures to investigate the
      evolution and reproduce observational properties of these
      systems. When considering the triple scenario, we determine the
      region of parameter space compatible with the observed binaries
      and use a publicly available machine-learning model to determine
      the dynamical stability of the triple system.}
  % results heading (mandatory)
{We find that, because of the extreme mass-ratio between binary components,
     tides have a limited impact, whatever the prescription used, and that
    the observed short orbital periods are at odds with post-mass-transfer
    scenarios.
   We also find that the overwhelming majority of triple systems compatible with the observed binaries are dynamically unstable and would disrupt within years of formation, forcing a hypothetical merger to happen so close to a zero-age main-sequence that it can be considered part of the star formation process.}
% conclusions heading (mandatory)
 {The most likely scenario to form such young, rapidly rotating,
    and short-period binaries is primordial rotation, implying that the
    observed binaries are pre-interaction. Our simulations further indicate
  that such systems will subsequently go through a common envelope and
  likely merge. These binaries show that the initial spin distribution of massive
  stars can have a wide range of rotational velocities.}
{}

\keywords{stars: rotation -- stars: massive -- binaries: close -- methods: numerical}

\maketitle
%
\section{Introduction}

Stellar rotation is a ubiquitous though as-yet poorly understood
phenomenon, especially for massive stars \cite[e.g.,][]{Langer_2012}.
Rotation is crucial for nucleosynthesis of certain elements
\citep[e.g., ``s-process''][]{limongi:18}, for powering transients
such as long gamma-ray bursts in the ``collapsar'' scenario
\citep[e.g.,][]{macfadyen:99}, and to understand the spins of neutron
stars and black holes \citep[e.g.,][]{callister:21, GWTC3}.
Understanding the rotational evolution of stellar surfaces and
interiors is thus important for galactic evolution, time domain astrophysics, and
gravitational-wave observations. Before investigating the origin of stellar
rotation, let us focus on how the rotation is
studied and what affects it.

\subsection{Observational constraints on stellar rotation}

For O- and B-type stars, spectroscopic observations typically provide only the \emph{projected}
  equatorial rotational velocity \vsini, that is the equatorial rotational
  velocity $v$ times the sine of the inclination angle to the line of
  sight $i$, which is usually unknown.

Usually, \vsini~ is measured through the Doppler broadening of spectral
lines, where the rotational component needs to be carefully separated
from other line-profile broadening components (e.g., macroturbulence, \citealt{Simon-Diaz2014}).

While investigating various stellar populations, common features in
the distribution of \vsini~ are found for O-type stars
\citep{Conti_1977, vfts_2013_otype, vfts_2015_otype}, B-type
stars \citep{Wolff_1982, Dufton_2013} and A-type stars \citep{Abt_1973}.
Indeed, the distribution of \vsini~ seems to peak near 100~\kms~ with
a tail of so-called fast-rotating stars with \vsini\,$>$\,200~\kms.

Recent studies describing large surveys (including stellar
populations in the Magellanic Clouds) have confirmed this overall
morphology of the \vsini~distribution of early-type massive stars
\citep[e.g.,][]{Dufton_2013,vfts_2013_otype,Holgado_2022}. Notably,
the spin distribution appears independent of the environment
\citep[e.g., Galactic field vs. young stellar
clusters,][]{Wolff_2007,Penny_2009, Huang_2010}. However, field stars
are on average older than stars in young open clusters, and since
rotation slows down as the star evolves because of stellar winds and the evolutionary expansion of the envelope, stars in young open cluster appear to rotate
systematically faster than stars in the field.
Interestingly, in more metal-poor environments, there is a trend
towards increasing mean rotational velocity in OB stars while keeping
the same overall non-uniform distribution \citep{Ramachandran_2019}. This is caused by the
  weakening of stellar winds at low metallicity
\citep[e.g.,][]{Vink_2001,mokiem:07}: the initial angular momentum of
the star is preserved longer. These results tentatively imply that the
observed \vsini~ distribution exists in environments of differing metallicity
and age.

The shape of the surface rotational distribution of massive
  stars is determined by various poorly understood processes, such as the unknown pre-main sequence (pre-MS)
  initial distribution and the influence of various physical processes
  active as the stars evolve and possibly interact with commonly
  occurring \citep[e.g.,][]{Sana_2012, moe_2017, Offner_2022} stellar
  companions \citep[e.g.,][]{packet:81, deMink_2013, vfts_2015_otype}.
This non-uniform (also referred as bimodal) rotational distribution is causing the main-sequence split which is observed on the color magnitude diagrams of the young star clusters \citep[see, e.g.][]{Kamann_2023,Wang_2023}.

\subsection{What affects the stellar rotation?}

The main factors that can affect the stellar rotation are: (\emph{i}) the
angular momentum budget left behind by star formation processes;
(\emph{ii}) the acting stellar physics processes throughout the
  evolution elapsed, e.g., magnetic braking, mass loss, changes in internal structure;
(\emph{iii}) binary interactions, including stellar mergers.

The initial spin of a star is the direct product of star formation
processes which happen embedded in opaque clouds. Different physical
aspects have been suggested to play a major role, such as magnetic
field or the presence of a circumstellar disc through the pre-MS stage
\citep[e.g.,][]{Rosen_2012}. In fact, these two factors are
proposed to be dominant in producing the non-uniform distribution
of rotation among the pre-MS low- and intermediate-mass stars
\citep{Bastian_2020}. This initial rotation distribution has not been
studied in the pre-MS massive star domain, but based on the similarity of observed \vsini~ distribution across spectral types, we can expect that a similar
 non-uniform initial distribution could also be present there.

After formation, the spin evolution depends on highly debated plasma physics (e.g., shears, magnetic
instabilities, see e.g., \citealt{Tayler_1973, spruit:99, spruit:02,Maeder_2000, fuller:19, denhartogh:20,ji:23}), uncertain wind mass
and angular momentum loss rates \citep[e.g.,][]{smith:14, renzo:17},
and the evolution of the stellar moment of inertia \citep[e.g.,][]{langer:98,zhao:20}.
Each of these evolutionary processes can modify the shape of the observed distribution of surface rotation \citep{Maeder_2000}.

Uncertainties affecting these stellar processes limit our
understanding of the angular momentum budget of a star in late evolutionary phases.
At core-collapse, hydrodynamical instabilities can redistribute this angular momentum between the newly formed compact object and the ejecta \citep[e.g.,][]{Kazeroni_2017}, complicating further the
connection between observed compact object spins and their parent stars' rotation.

In addition, the vast majority of massive stars exist in multiple systems \citep[e.g.,][]{Mason_2009, Duchene_2013, Sana_2013, Kobulnicky_2014, Dunstall_2015, Almeida_2017, Offner_2022}.
In such systems, interactions among stars are expected to occur in 70 \% of cases during the main-sequence (MS) stage evolutionary stage \citep{Sana_2012}.
Such binary interactions may also play a dominant role in shaping the observed
distribution of surface rotation rate, especially for the tail of fast rotators because of tides, mass transfer, or mergers. This idea has been proposed
  multiple times both from a theoretical perspective
  \citep{packet:81, Pols_1994, deMink_2013, vinciguerra:20} and observational studies \citep{blaauw:93,  vfts_2015_otype, Cazorla2017b, Britavskiy_2023}.
One example is the famous category of Be stars, which are fast rotating stars with decretion disks \citep{Fremat_2005, Rivinius_2013}.
They are not known to be paired with main-sequence companions \citep{Bodensteiner_2020}, but instead, many of them are orbited by a stripped star \citep{Wang_2018,Wang_2021} or a compact companion \citep[e.g.,][]{Reig_2011,Dodd_2023}.

However, even if binarity causes the fast-rotating tail of the distribution, this still leaves the bulk of the distribution unexplained. For example, \citet{vfts_2015_otype} suggested that
the initial rotation distribution of massive stars is the same for
apparently single and single-lined spectroscopic binary (SB1) stars.
This gives us a hint that some of the rapid rotators can be effectively single or still be in the pre-interacting binary regime.
Conversely, some of the young spectroscopic binaries are still in the pre-interacting stage and the rotation of its components has not yet been affected by binary interaction. A few examples have been proposed in the past \citep{Johnston2021,Stassun_2021, Naze_2023_rot}.
To distinguish what exactly causes the fast rotation, we need to investigate in detail different binary configurations and their evolution from the rotation point of view.

To summarize, stellar rotation could indirectly indicate whether a binary interaction has happened in a given system at some point in its evolution.
To consider rotation as a tracer of binary interaction, it is however important to understand which factors affect it.

Answering these questions will help us to understand the overall evolution of stellar spin, from stellar birth to the latest various endpoints of stellar evolution.

\subsection{Main motivation of the present work}

Here, we study
  known Galactic fast-rotating OB stars that have been identified
  in single-lined spectroscopic binaries, specifically HD\,25631,
  HD\,191495, and HD\,46485
  \citep[SB1,][see also \Tabref{table:sb1_fast}]{Britavskiy_2023,Naze_2023_rot}. Taking into account
  their high surface rotational velocities (\vsini\,$>$\,200~\kms), short orbital periods ($P$ $<$ 7 days), and very unequal
  mass-ratios, we aim at understanding what causes such a high rotation, and what is the effect (if any)
of past and future binary interactions. Indeed, their observed \vsini~ is significantly larger than the tidal synchronization velocities.
These systems are not exceptional: similar short-period
binaries with an extreme mass ratio are known \citep[][]{Moe_2015,Jerzykiewicz_2021}. Thus, by evaluating different processes that
could result in a high rotation for the investigated targets we can
emphasize what is the more probable scenario to form the whole class of such binaries.
The main goal of the current work is to evaluate different evolutionary scenarios that could result in these peculiar binary configurations and predict how they will further evolve.

Numerical calculations of the internal evolution of both stars in an
interacting binary accounting for the exchange of mass and angular
momentum \citep[e.g.,][]{wang:20, Badry_2021,Renzo_2021, pauli:22,
  renzo:23} allow us to investigate the detailed evolution of
rotational velocities in such systems.

We follow the evolution of these targets up to late stages, in order
to understand whether they will be the sources of stripped stars, stellar mergers and/or X-ray binaries.

The paper is organized as follows. In \Secref{sec:sample}, we introduce the sample of fast rotators and the methodology we are following; from \Secref{sec:pre-interaction} to \Secref{sec:triple_scenario} we are evaluating three different scenarios of binary interaction that could occur in a given system.
The primordial rotation scenario (pre-interacting regime) is presented
in \Secref{sec:pre-interaction}. We evaluate how long a high
rotational velocity can remain assuming that these systems were formed
with initial rotation close to the current \vsini. In addition, we highlight what is the effect of tides on the rotation and which further binary interaction will affect such systems.
In \Secref{sec:post-interaction} we evaluate if the present orbital configuration of the three systems could form through binary interaction. We assume that the accretor has gained its high rotational velocity via stable mass transfer from the donor.
In \Secref{sec:triple_scenario} we present our discussion regarding merger in a triple system. A merging process could result in a fast rotating star, however, dynamic stability parameters that are able to reach the observed binary configurations need to be investigated. 
\Secref{sec:discussions} presents a general discussion, while \Secref{sec:conclusions} recalls the main conclusions of this paper.

\section{Sample of stars and the adopted methodology}
\label{sec:sample}

Recently, three peculiar short-period binaries with a fast-rotator
(\vsini\,$>$\,200~\kms) as primary component \citep{Naze_2023_rot}
were detected and analyzed in detail. These three targets (HD\,25631,
HD\,191495 and HD\,46485) have common morphology of light-, and
radial-velocity curves, namely two narrow eclipses and a strong
reflection effect through one orbital cycle indicating that the
secondary component has a lower surface temperature than its primary
and is thus \emph{not} a compact object. The orbital and
fundamental physical parameters of these targets have been studied in
detail in \citet{Naze_2023_rot} and are summarized in \Tabref{table:sb1_fast}. Notably, these three SB1 systems
share short orbital periods (3.2 to 6.9 days) and rather extreme
mass ratios ($q=M_2/M_1=M_\mathrm{unseen}/M_\mathrm{seen}$ from 0.035 to 0.15).

The position of primaries of the selected systems on the Hertzsprung--Russell (HR)
diagram is presented in \Figref{fig:hr_plot_paper2023} together
with a sample of 50 Galactic fast-rotating O stars from \citet{Britavskiy_2023} for reference.
In addition, we calculated the evolutionary tracks that represent the main-sequence evolutionary stage of a non-rotating single star at solar metallicity (Z = 0.02).
For each track, the zero-age main-sequence (ZAMS) was defined at the moment when total nuclear luminosity from all burning becomes equal to the surface luminosity within the absolute difference of $10^{-4}$.
The end of MS is defined as the central mass fraction of hydrogen decreasing below $X(^{1}\mathrm{H})<0.01$.

The ages for these systems were
derived by fitting the BONNSAI evolutionary tracks \citep{Bonnsai_2014} at Galactic
metallicity to the derived effective temperatures and luminosities of
the primaries. These are in agreement with the age estimates of the
parent clusters for two systems belonging to clusters, namely
HD\,46485 in NGC 2244 and HD\,191495 in NGC 6871 \citep[see for details][]{Naze_2023_rot}.

The existence of these systems raises questions regarding
their origin and evolutionary paths, which can shed light on the initial
\vsini~distribution of OB stars. The first question to address is: what is the origin of the super-synchronous rotation of the primaries in these systems?

\begin{table}[!htbp]
{\small
%\centering
\caption{Basic physical and orbital parameters of three
  fast-rotating SB1 systems that have been investigated by
  \citet{Naze_2023_rot}. }
\label{table:sb1_fast}
\begin{tabular}{l c c c}
\hline
\hline
Target                  & HD\,25631          &  HD\,191495      & HD\,46485 \\
\hline
SpC (primary)             &  B3V             &  B0V             &  O7\,V((f))nvar?  \\
log($L_{1}/L_{\odot}$)    &  3.24 $\pm$ 0.02 & 4.51 $\pm$ 0.04  & 4.95 $\pm$ 0.03 \\
T$_{eff,1}$ [kK]          & 19.9 $\pm$ 0.5$^{\ast}$  & 28.2 $\pm$ 1.3   & 36.1 $\pm$ 0.7  \\
\vsini~ [\kms]            &  221$\pm$ 10 &  201$\pm$ 10  & 334$\pm$ 16  \\
\hline
age  [Myr]       &  9.6$\pm$9.4  &  7.6$\pm$1.8  & 2.5$\pm$1.1   \\
$M_\mathrm{1}$   [$M_{\odot}$]  &  7 -- 8  &  15    & 24  \\
$M_\mathrm{2}$   [$M_{\odot}$] &  0.8 -- 1.2 & 1.2 -- 1.8 &  0.8 -- 1.1 \\
$P$  [d]  &  5.2  & 3.6   &  6.9 \\
$e$       & 0   &  0  &  0.033 \\
$i$ [$\deg$]  &  79 -- 80  &  67 -- 69  &  73 -- 76 \\
$q$ & 0.12 -- 0.15 & 0.08 -- 0.12 & 0.035 -- 0.045 \\
\hline
\end{tabular}
Notes. $^{\ast}$ - the uncertainty in T$_{eff}$ taken as average uncertainty of the method presented in \citet{Braganca_2012}.
}
\end{table}

\begin{figure}[!ht]
  \centering
  \script{hr_plot_paper2023_fin.py}
  \includegraphics[width=0.50\textwidth]{HR_REAL_ROT_Paper2023.pdf}
  \caption{Location in the HR diagram of the Galactic OB fast rotators studied in \citet{Britavskiy_2023} and \citet{Naze_2023_rot}. Apparently single, binary and runaway fast-rotators are marked by specific symbols. Non-rotating evolutionary tracks represent the main-sequence stage of a single-star evolution at solar metallicity for different initial masses.}
  \label{fig:hr_plot_paper2023}
\end{figure}

To model the different evolutionary scenarios we used the MESA
\citep[Modules for Experiments in Stellar Astrophysics, version
15140][]{Paxton2011, Paxton2013, Paxton2015, Paxton2018, Paxton2019,
  Jermyn2023}. The adopted input parameters and output results for our
models are available on GitHub\footnote{\url{https://github.com/NikolayBritavskiyAstro/fast_rotating_binaries}} and Zenodo\footnote{\url{https://zenodo.org/records/10479754}}. As a
starting point we took the setup from \citet{Renzo_2021} based
  on reproducing the Galactic O-type fast rotator $\zeta$ Ophiuchi.

All our models have a solar metallicity Z = 0.02, considering that the observed sample consists of
Galactic stars. In the subsequent sections, we describe the main
physical ingredients relevant to the specific scenario we consider,
\Appref{sec:MESA_setup} describes the full list of physical and
numerical assumptions. Following the "best practice" tips of working with MESA, we repeated calculations by performing resolution tests (see \Appref{sec:resolution_test} for details).

\section{Primordial rotation -- Pre-interaction scenario}
\label{sec:pre-interaction}

In this section, we assume that the primaries of HD\,25631, HD\,191495, and HD\,46485 were
born with a high rotation and we describe their subsequent
evolution.
Our main question is how long such a fast initial rotation can remain - i.e. could such initial rotation be detected in practice -  and what is the effect of tides in such short-period systems?

\subsection{Method}

As we are investigating the surface rotational properties of massive stars, it is very important to take into account the main factors that affect it. In the case of single stars, we need to consider assumptions regarding the
mass loss through stellar winds, core overshooting, and expansion of the envelope while the star is evolving. For the consideration of mass loss, we use the \texttt{'Dutch'}
wind scheme that combines \citet{Vink_2001} mass loss prescription,
suitable for effective temperatures $\geq$ 11 000 K and mass fraction
of hydrogen $X(^{1}\mathrm{H})$ $>$ 0.4, with \citet{Jager_1988}
mass loss rates for temperatures lower than 10 000 K. In between the
mentioned temperatures, the \texttt{'Dutch'} scheme is linearly
interpolating hot and cold wind prescriptions.
To test robustness of our conclusions against uncertainties in wind mass loss rates \citep{renzo:17}, we also compute models with the "flat" wind mass loss rate from \citet{Bjorklund_2023}.
 Regarding extended convective mixing, we use the exponential core overshooting based on
\citet{Herwig_2000} with free parameters chosen following
\cite{claret:18} work which reproduces the width of the main-sequence in 30
Doradus predicted by \citet{Brott_2011}. These adopted parameters are essential for a proper definition of stellar
structure (e.g., mass of the core, mass loss, etc) taking into account
that they will affect the rotational properties of the stars under
investigation.
While investigating the evolution of surface rotational velocity, it is important to be aware of which prescription of angular momentum (AM) transport has been adopted, in all our models we assume a Spruit - Tayler dynamo mechanism \citep{spruit:02}.
To test how it affects the surface rotation we ran several models without it and in addition, we ran the prescription of \citet{fuller:19} AM transport mechanism that is more efficient with respect to a Spruit - Tayler dynamo.

\subsection{Results and interpretation (effect of mass loss)}
\label{sec:res_primordial}

We start our modeling from the investigation of the rotational
properties of a 24\,$M_{\odot}$ star evolving as a single star. First, we
show for illustrative purposes in \Figref{fig:v_surf_one_star} the MS
evolution of the surface equatorial rotational velocity
($v_\mathrm{rot}$) with varying initial values
$v_\mathrm{rot}^\mathrm{ini}$ at solar metallicity ($Z=0.02$).

Our choice of initial mass roughly corresponds to the primary
  mass of HD\,46485 which is the largest one in our sample, thus the one
most affected by wind spin-down. Indeed, in
\Figref{fig:v_surf_one_star} we can see that, regardless of the
initial rotation, the star slows down to
$v_\mathrm{rot}\sim 5$\,\kms~ within $\sim$8\,Myrs \citep[solid lines,][]{Vink_2001}. The evolutionary tracks stop at the
terminal age main-sequence (TAMS, defined by the central mass fraction
hydrogen decreasing below of $X(^{1}\mathrm{H})<0.01$). We can
notice around $\sim7$\,Myr the large decrease in rotation and mass caused by
  the increase in mass loss rate due to the ``bi-stability jump''
(e.g.,\citealt{Vink_2001} but see also \citealt{bjorklund:21})
which also removes a large portion of the available angular
momentum of the star. The physical reason that causes
such significant changes in mass loss rate is the recombination of
  iron ions at certain effective temperatures ($\sim$ 21 000 and 10
000 K) which dramatically increases the opacity and thus the
  effectiveness of the line-driven wind in removing mass, and as a result it reduces the angular momentum.
This rotation braking  occurs mostly for O-type stars at late stages of the MS. However, the stellar wind model based on \citet{Bjorklund_2023} study does not predict the existence of the ``bi-stability jump'' (see dashed lines in \Figref{fig:v_surf_one_star}) because it is based on a different radiation-driven mass loss prescriptions.

\begin{figure}[!ht]
  \centering
  \script{plot_save_mesa_comparison_wind_fin.py}
  \includegraphics[width=0.50\textwidth]{age_vsurf_comparison_onestar.pdf}
  \caption{Evolution of equatorial surface rotational velocity of a
    single 24 $M_{\odot}$ star for different values of initial
    rotational velocity by assuming the formalism of \citet{Vink_2001} stellar wind model (solid lines) and \citet{Bjorklund_2023} wind model (dashed lines). In addition, tracks without the Spruit - Tayler dynamo mechanism (dotted-dashed lines) and the tracks assumed \citet{fuller:19} AM transport (dotted lines) are added for reference.
 The tracks represent only the main-sequence
    stage up to terminal-age evolutionary phase. The error bars
    represent the observed position of HD\,46485, whose primary has comparable mass.}
  \label{fig:v_surf_one_star}
\end{figure}

Thus, there is still no general consensus regarding
which mass loss prescriptions \citep{smith:14, renzo:17,
  Keszthelyi_2017, bjorklund:21, Bjorklund_2023} and which model of
internal angular momentum transport \citep[e.g.,][]{fuller:19,
  denhartogh:20, Gormaz_2023} are more suitable for massive OB stars.
Consequently, there is a systematic theoretical uncertainty
  in the steepness of $v_\mathrm{rot}$ drop and
the size and duration of the bi-stability jump can vary depending on
which underlying physics is adopted (e.g., mixing efficiencies,
mass loss rates, and magnetic braking). Another effect
  contributing to the spin down of the star is the evolutionary
expansion of the envelope, controlled primarily by the change in
  mean molecular weight \citep[e.g.,][]{xin:22} and thus
  what fraction of the hydrogen mass is burned in the convective
  core and ultimately the convective boundary mixing \cite[e.g.,][]{Brott_2011, johnston:21}.

Because of conservation of AM, as the radius increases the surface rotational frequency has to decrease -- with angular momentum transport and wind losses adding complications to this simplified argument.
However, since all our targets are located close to ZAMS, there is little time for this to happen.
Without the Spruit - Tayler dynamo mechanism, AM transport less efficient, thus the stars spin down faster because of the wind (see dashed-doted lines in \Figref{fig:v_surf_one_star}).
Conversely, assuming the more efficient AM transport from \citet{fuller:19}, the star is spin down more slowly (see dotted lines in \Figref{fig:v_surf_one_star}).
This illustrates that while a star is evolving, the core contracts and spins up and the resulting surface spin down rate directly depends on the assumed efficiency of AM transport.
However, the difference in surface rotational velocity caused by different AM transport mechanisms is small within the early MS stage.

Another factor that influences the rotation rate is the assumed wind mass loss rate.
A good illustration of how wind underlying theory is affecting the evolution of $v_\mathrm{rot}$ can be seen comparing tracks computed with \citet{Bjorklund_2023} (dashed), where the increase in wind mass loss because of bi-stability jump(s) is absent, and those computed with \citet{Vink_2001} (solid lines in \Figref{fig:v_surf_one_star}).
The overall mass loss rate in the former is lower, allowing the stars to retain more AM and a faster surface rotation rate throughout the main sequence.
The difference between \citeauthor{Bjorklund_2023} and \citeauthor{Vink_2001} mass loss rates lies in the new prescription of radiative acceleration which covers a large range of atmosphere layers including supersonic wind outflow which is considered in \citet{Bjorklund_2023} approach.
The mass loss rate is then derived as a scaling factor of fundamental stellar parameters based on a large grid of models.
The resulting mass loss rate appears to be weaker than the previous estimates and as a result the $v_\mathrm{rot}^\mathrm{ini}$ may remain high for a significant time (up to the end of the MS).

However, differences in $v_\mathrm{rot}$
behavior become significant only near the end of the MS, a stage far from
current evolution properties of our targets (see \Figref{fig:hr_plot_paper2023}). The results we presented in
\Figref{fig:v_surf_one_star} can therefore be adopted as a typical
$v_\mathrm{rot}$ behavior for the young OB star domain. The resolution
test for a given star model is presented in
\Figref{fig:resolution_test_one_star}.

\subsection{Tidal interaction}

The next step is to consider stars evolving in a binary system and add the effect of tides on the rotational velocity
and how they affect the evolution of the surface rotational
  velocity of our systems, shown in \Figref{fig:v_surf}. We
start with binaries with architectures comparable to the ones observed
today, with fast-rotating super-synchronous primaries and initial rotation of secondaries set to $v_\mathrm{rot}^\mathrm{2,ini}$ = 1 \kms.
We run three binary systems with ($M_\mathrm{1,ini}$, $M_\mathrm{2,ini}$,
$P_\mathrm{ini}$, $v_\mathrm{rot}^\mathrm{1,ini}$ ) = (24 $M_{\odot}$, 1 $M_{\odot}$, 6.9 d, 350 \kms), (15 $M_{\odot}$, 1.5 $M_{\odot}$,  3.6 d, 200 \kms), (7 $M_{\odot}$, 1 $M_{\odot}$, 5.2 d, 220 \kms) for HD\,46485, HD\,191495, HD\,25631, respectively (based on the values presented in \Tabref{table:sb1_fast}).
By default we used the \citet{Vink_2001} mass loss prescription and we evolved the systems beyond the initial onset of mass transfer.

To explore the large uncertainties in modeling of tidal interactions
\citep[e.g.,][]{zahn:75, Zahn_1977, Qin_2018, preece:22, fuller:22}, we use three different
available tidal synchronization algorithms. Namely, we used
no tides at all (red dot-dashed lines in \Figref{fig:v_surf}), the tidal synchronization prescription for radiative envelopes
(\texttt{sync\_type = "Hut\_rad"}, aka "MESA default", see black solid lines) based on \citet{Hut_1980} and
\citet{Hurley_2002} as well as the structure-dependent
prescription (\texttt{sync\_type = "structure\_dependent"})
implemented in the POSYDON code \citep[blue dashed lines, based on][]{Qin_2018, Fragos_2023}.

\begin{figure}[!ht]
  \centering
  \script{plot_save_mesa_comparison_panels_fin.py}
  \includegraphics[width=0.50\textwidth]{age_vsurf_comparison_panel.pdf}
  \caption{Evolution of equatorial and synchronization surface rotational velocity for our
    three program systems by taking into account different tidal
    synchronization prescriptions. "MESA default" refers to the
    radiative envelope tidal synchronization (\texttt{"Hut$\_$rad"}
    prescription), and "POSYDON" to the structure dependent
    prescription. We emphasize that different scales for the axes of
      each panel are used due to the different rotation rates, stellar
      masses, and thus lifetimes of the systems.}
  \label{fig:v_surf}

\end{figure}

\begin{figure}[!ht]
  \centering
  \script{plot_save_mesa_comparison_panels_acc_fin.py}
  \includegraphics[width=0.50\textwidth]{age_vsurf_acceleration_panel.pdf}
  \caption{The same as \Figref{fig:v_surf} but assuming the $v_\mathrm{rot}^\mathrm{1,ini}$ = 1 \kms. The gray dashed lines represent the evolution of synchronization rotational velocity for each system.}
  \label{fig:v_surf_acceleration}
\end{figure}

The main difference between these prescriptions is the different treatment of the synchronization timescale for radiative and convective layers depending on the
current evolutionary stage of the star.
 In POSYDON, the structure-dependent prescription checks the shortest synchronization timescale between the dynamical tides (same as \texttt{sync\_type = "Hut\_rad"}) and equilibrium tides based on the existing convective layers (similar \texttt{sync\_type = "Hut\_conv"} but not limited to a deep convective envelope) and applies it for each layer of the star, while in "MESA default" for each layer of the star the code takes into account only one dynamical timescale (\texttt{sync\_type = "Hut\_rad"}).
It is important because the mechanism of the dissipation processes of the tidal kinetic energy is different for the convective and radiative envelopes (equilibrium and dynamical tides respectively).
 Also, the code takes into account the different distribution of mass in the star depending on these
layers.
